---
title: "Proportional Hazards Model slides "
#subtitle: "This is a Slide Template"
author: "Hermela Shimelis"
date: last-modified
date-format: long
format:
  revealjs:
    transition: fade
    css: custom.css
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
csl: ieee.csl
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---



## Cox Proportional Hazards Model {.smaller}

-   The Cox proportional hazards model is used to estimate the time until a particular event occurs, including death or disease [@Tibshirani2022-yr] [@Bellera2010-fq].

-   Assesses the effect of covariates on the hazard or risk given by a hazard ratio

-   Cox regression model is based on the hazard function $h_x(t)$ with covariates at time t given by: 

$$h_x(t)=h_0(t)\exp(\beta_1x_1 +\beta_2x_2 + \dots + \beta_p x_n) $$
$$h_x(t) = hazard \ function\ $$ 
$$h_0(t) = Baseline \ hazard \ function\ $$
$$\beta_1x_1 +\beta_2x_2 + \dots + \beta_p x_n = linear\ combination\ of \ covariates\ and\ their\ coefficients$$ 

## Assumptions of Cox regression model

-   Proportionality of the hazards: the factors investigated have a constant impact on the hazard - or risk - over time [@Bellera2010-fq].

- Graphical strategies to assess proportionality assumption

  - Kaplan-Meier Survival Distribution $S(t)$: Plot $S(t)$ as a function of survival time for each level covariate.

  - plot the function $log(-log(S(t)))$ as a function of the log survival time (log represent natural log).
  
  - Schoenfeld Residuals
  
- Consider using different approaches if assumption is not met

## Survival data visualization

-    E.g., Kaplan-Miers plot of the lung dataset in R

```{r, warning=FALSE, echo=F, message=FALSE}
# Load necessary libraries
library(survival)
library(survminer)

# Load the lung dataset
data(lung)

# Fit a Kaplan-Meier survival curve
fit <- survfit(Surv(time, status) ~ sex, data = lung)
```


```{r, warning=FALSE, echo=F}
# Plot the Kaplan-Meier survival curves
ggsurvplot(fit, data = lung, 
           pval = TRUE, 
           conf.int = TRUE, 
           risk.table = TRUE, 
           ggtheme = theme_minimal(), 
           legend.title = "Sex",
           legend.labs = c("Male", "Female"))

# Display the plot
#print(ggsurvplot)

```




## Hazard ratio from Cox regression model

-    E.g., Hazard ratio of the lung dataset in R

```{r, warning=FALSE, echo=F, message=FALSE}
# Run the Cox proportional hazards model
cox_model <- coxph(Surv(time, status) ~ age + sex + ph.ecog, data = lung)

# Display the summary of the Cox model to get the coefficients and other details
#summary(cox_model)

# Get the exponentiated coefficients (hazard ratios) and their confidence intervals
hazard_ratios <- exp(coef(cox_model))
conf_int <- exp(confint(cox_model))

# Create a data frame to display the hazard ratios and confidence intervals clearly
hazard_ratios_df <- data.frame(
  #Variable = names(hazard_ratios),
  HazardRatio = round(hazard_ratios,2),
  Lower95CI = round(conf_int[,1],2),
  Upper95CI = round(conf_int[,2],2)
)


```

```{r, warning=FALSE, echo=F}
library(knitr)
kable(hazard_ratios_df)
```
## References


